Resonant Scattering in a Uniform Sphere with Large Optical Depth

### 2021-09-14 ms.tex
Fixing section in introduction describing Harrington. Removed duplicated info, made description of errors more direct

Numerical simulations can be used to understand the complex layers of these irradiated exoplanet atmospheres. However, the extreme optical depths at \lya line center introduce a computational barrier to simulating radiative transfer via Monte Carlo methods. The main impediment to performing magnetohydrodynamic simulations with fully a coupled radiative transfer solution via Monte Carlo methods is the prohibitive computational cost. The number of scatterings a photon undergoes is proportional to the line center optical depth.  Near the base of the atmosphere, the line center optical depth is ${\sim}10^9$, so most of the time is spent integrating photons in these cells. A method that can treat these zones via a simple solution has the potential to greatly accelerate the calculation. A complete, general analytic solution for the resonance scattering has never been derived but approximate solutions exist in certain limits.

\citet{1973MNRAS.162...43H, 1974MNRAS.166..373H} demonstrated that resonance-line radiation transfer in low-density, optically-thick media can be described by the Poisson equation. However, the accuracy of their solution is limited by the treatment of the boundary conditions. In their eigenfunction expansion, the spatial and frequency components of the differential equation are separated, yet the spatial eigenvalues are defined as being explicitly frequency-dependent. The resulting solution is therefore invalid on the basis that the differential equation and boundary conditions cannot simultaneously be satisfied. A more detailed discussion of the boundary condition is included in Section \ref{sec:steadystate}. \citet{1990ApJ...350..216N} follows this approach in an attempt to extend the result to media of intermediate optical depth, including the effects of scattering in the Doppler core of the line, but again does not use the correct boundary condition. This produces a divergent, inhomogeneous solution to the problem in 1D slab geometry. \citet{2006ApJ...649...14D} generalize the problem to spherical geometry, which is the same physical set up used here. Following this, \citet{2020arXiv200509692L} presents closed-form solutions for the radiation field in spherical geometry, utilizing the ``core skipping'' method discussed in \citet{2015MNRAS.449.4336S} to handle photon frequency redistribution within the Doppler core. This technique involves drawing the atom's velocity from a truncated Gaussian distribution, forcing photons back into the wing when they scatter into the core. This is motivated by the assumption that all core scattering results in zero spatial diffusion until the photon is redistributed back out into the wing where the mean free path becomes much larger and it can diffuse in space again. However, it is not clear whether this prescription is fully independent of the choice of velocity distribution.

The question is how to satisfy the surface boundary condition in Eq. \ref{eq:bc}. On this point, we note that the derivation in \citet{1973MNRAS.162...43H}, and numerous studies later studies, does not actually produce a valid solution of the transfer equation and boundary condition. The reason is as follows. In the notation of \citet{1973MNRAS.162...43H}, a separation of variables
$J(\tau,\sigma) = \theta(\tau) j(\sigma)$ in spatial variable $\tau$ and frequency variable $\sigma$ is attempted (their Equations 16 and 23). The solutions for the eigenfunctions $\theta(\tau)$ and $j(\sigma)$ then depend explicitly on the separation constant $\lambda$. In order to satisfy the boundary conditions, the separation constant must satisfy an eigenvalue equation of the form
\be
\lambda \tan(\lambda B) & = & \frac{3}{2} \phi,
\label{eq:evalue}
\ee
where $2B$ is the slab optical depth at line center, and $\phi$ is the line profile ($\phi(\nu)\Delta$ here). The key point is that the line profile depends on one of the coordinates, frequency, and this causes the eigenvalues of the separation constant to depend on frequency. But then the separation constant is not constant, and it does not satisfy Eq. \ref{eq:rt_no_destr} since the frequency derivatives will act on the separation ``constant", giving extra terms. To our knowledge the errors introduced by this ansatz have never been quantified.

In the limit of large optical depth $B$, \citealt{1973MNRAS.162...43H} approximates the eigenvalues as $\lambda_n B \simeq \pi (n-1/2)$, which gives $J=0$ at the surface. This is essentially the result in Eq. \ref{eq:J0} above. However, since this solution gives zero at the surface, their Eq. 34 subsequently allowed the separation constant to have a small deviation from the above expression, which was explicitly frequency dependent. This allowed a nonzero intensity at the surface, but at the cost of rendering the separation of variables assumption invalid.

>>>>

Numerical simulations can be used to understand the complex layers of these irradiated exoplanet atmospheres. However, the extreme optical depths at \lya line center introduce a computational barrier to simulating radiative transfer via Monte Carlo methods. The main impediment to performing magnetohydrodynamic simulations with fully a coupled radiative transfer solution via Monte Carlo methods is the prohibitive computational cost. The number of scatterings a photon undergoes is proportional to the line center optical depth.  Near the base of the atmosphere, the line center optical depth is ${\sim}10^9$, so most of the time is spent following photons in these cells. A method that can treat these zones via a simple solution rather than Monte Carlo has the potential to greatly accelerate the calculation. A complete, general analytic solution for the resonance scattering has never been derived but approximate solutions exist in certain limits.

\citet{1973MNRAS.162...43H} demonstrated in slab geometry that resonance-line radiation transfer in optically-thick media can be described by the Poisson equation. However, their solution contains an incorrect treatment of the boundary condition. To our knowledge, the errors introduced by this treatment have never been quantified. They attempt a separation of variables $J(\tau,\sigma) = \theta(\tau) j(\sigma)$ in spatial variable $\tau$ and frequency variable $\sigma$ (their Equations 16 and 23). The solutions for the eigenfunctions $\theta(\tau)$ and $j(\sigma)$ then depend explicitly on the separation constant $\lambda$. In order to satisfy the boundary conditions, the separation constant must satisfy an eigenvalue equation of the form
\be
\lambda \tan(\lambda B) & = & \frac{3}{2} \phi \Delta,
\label{eq:evalue}
\ee
where $2B$ is the slab optical depth at line center, and $\phi$ is the line profile. The key point is that the line profile depends on one of the coordinates, frequency, and this causes the eigenvalues of the separation constant to depend on frequency. Thus, the separation constant is not constant, and it does not satisfy the Poisson equation since the frequency derivatives will act on the separation ``constant", giving extra terms. In the limit of large optical depth $B$, they approximate the eigenvalues as $\lambda_n B \simeq \pi (n-1/2)$. However, this solution gives zero mean intensity at the surface. Their Eq. 34 subsequently allowed the separation constant to have a small deviation from the above expression, which was explicitly frequency dependent. This allowed a nonzero intensity at the surface, but at the cost of rendering the separation of variables assumption invalid. Our treatment, using the correct boundary condition, quantifies the errors in their ansatz.

\citet{1990ApJ...350..216N} follows this approach in an attempt to extend the result to media of intermediate optical depth, including the effects of scattering in the Doppler core of the line, but again does not use the correct boundary condition. This produces a divergent, inhomogeneous solution to the problem in 1D slab geometry. \citet{2006ApJ...649...14D} generalize the problem to spherical geometry, which is the same physical set up used here. Following this, \citet{2020arXiv200509692L} presents closed-form solutions for the radiation field in spherical geometry, utilizing the ``core skipping'' method discussed in \citet{2015MNRAS.449.4336S} to handle photon frequency redistribution within the Doppler core. This technique involves drawing the atom's velocity from a truncated Gaussian distribution, forcing photons back into the wing when they scatter into the core. This is motivated by the assumption that all core scattering results in zero spatial diffusion until the photon is redistributed back out into the wing where the mean free path becomes much larger and it can diffuse in space again. However, it is not clear whether this prescription is fully independent of the choice of velocity distribution.
